Accuracy in Parameter Estimation and Simulation Approaches for Sample Size Planning Accounting for Item Effects

### 2024-12-01 manuscript_draft.Rmd
fixing figure labels and referencees

1)  How does pilot data influence sample size suggestions?

@@

```{r decile50-figure, fig.cap = "Comparison of the cutoffs for 5th decile across heterogeneity (columns), powering of items (rows), and scale size (color).", fig.height=8, fig.width=8}
summary_long %>% 
  filter(name == "Decile 50") %>% 
  group_by(variability, original_n, scale_size, name, power) %>% 
  summarize(new_sample = mean(new_sample)) %>% 
  ggplot(., aes(original_n, new_sample, color = scale_size)) + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line(aes(original_n, original_n), color = "black") + 
  geom_point() + 
  geom_line() +
  theme_classic(base_size = 10) + 
  facet_grid(power~ variability,
             labeller = labeller(power = power.labs)) + 
  ggtitle("Results for 5th decile") +
  xlab("\nPilot Sample Size") + 
  ylab("Corrected Sample Size\n") + 
  scale_shape_discrete(name = "Scale Size") +
  scale_color_manual(name = "Cutoff Decile",
                     values = c(plot_colors)) + 
  theme_bw() +
  theme(legend.position="bottom")

@@

```{r bimodal-figure, fig.cap = "Simulated example of Likert data with half of the items as bimodal distributions.", fig.height=10, fig.width=7, warning = F, include = F, message = F}
ggplot(researcher_sample %>% 
         filter(item < 11), aes(x = score)) +
  geom_histogram(aes(y = ..count..), bins = 7, position = "identity", alpha = 0.7, color = "black") +  # Stacked histograms
  labs(x = "Score", y = "Item") +
  theme_minimal() +
  facet_wrap(~factor(item), scales = "free_y", nrow = 10) +  # Place each item in a separate row
  theme(legend.position = "none",
        axis.text.y = element_blank()) + 
  coord_cartesian(xlim = c(1, 7))  # Set x-axis limits
```

>>>>

1)  How do pilot data influence sample size suggestions?

@@

```{r decile50-figure, fig.cap = "Comparison of the cutoffs for 5th deciles across heterogeneity (columns), powering of items (rows), and scale size (color).", fig.height=8, fig.width=8}
summary_long %>% 
  filter(name == "Decile 50") %>% 
  group_by(variability, original_n, scale_size, name, power) %>% 
  summarize(new_sample = mean(new_sample)) %>% 
  ggplot(., aes(original_n, new_sample, color = scale_size)) + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line(aes(original_n, original_n), color = "black") + 
  geom_point() + 
  geom_line() +
  theme_classic(base_size = 10) + 
  facet_grid(power~ variability,
             labeller = labeller(power = power.labs)) + 
  ggtitle("Results for 5th deciles") +
  xlab("\nPilot Sample Size") + 
  ylab("Corrected Sample Size\n") + 
  scale_shape_discrete(name = "Scale Size") +
  scale_color_manual(name = "Cutoff Decile",
                     values = c(plot_colors)) + 
  theme_bw() +
  theme(legend.position="bottom")

@@

```{r bimodal-figure, fig.cap = "Simulated example of Likert data with half of the items as bimodal distributions.", fig.height=10, fig.width=7, warning = F, message = F}
ggplot(researcher_sample %>% 
         filter(item < 11), aes(x = score)) +
  geom_histogram(aes(y = ..count..), bins = 7, position = "identity", alpha = 0.7, color = "black") +  # Stacked histograms
  labs(x = "Score", y = "Item") +
  theme_minimal() +
  facet_wrap(~factor(item), scales = "free_y", nrow = 10) +  # Place each item in a separate row
  theme(legend.position = "none",
        axis.text.y = element_blank()) + 
  coord_cartesian(xlim = c(1, 7))  # Set x-axis limits
```

### 2024-12-01 manuscript_draft.tex
fixing figure labels and referencees

\emph{Step 1}. The concreteness ratings data includes 27031 concepts that were rated for their concreteness. We randomly selected \emph{n} = 20 abstract words (\(M_{Rating}\) \textless= 2) and \emph{n} = 20 concrete words (\(M_{Rating}\) \textgreater= 4). In the original study, not every participant rated every word, which created uneven sample sizes for each word. Further, participants were allowed to indicate they did not know a word, and those responses were set to missing data. In our sample of 40 words, the average pilot sample size was 28.52 (\emph{SD} = 1.80), and we will use 29 as our pilot sample size for the concreteness ratings (this information will be used in the follow-up to the simulation study).

We first selected the same real words in the ELP data as the concreteness subset selected above, and this data includes 27031 real words. The average pilot sample size for this random sample was 32.67 (\emph{SD} = 0.57), and \emph{n} = 33 will be our pilot size for the lexical decision task.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  How does pilot data influence sample size suggestions?
\end{enumerate}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{manuscript_draft_files/figure-latex/decile50-figure-1.pdf}}
\caption{\label{fig:decile50-figure}Comparison of the cutoffs for 5th decile across heterogeneity (columns), powering of items (rows), and scale size (color).}
\end{figure}

@@

The above simulations assume symmetric population distributions with small, medium, or large heterogeneity or a skewed distribution with the same heterogeneity. However, as shown in Pollock (2018), sample rating data that appears to have a mean distribution may represent the average two separate distributions (i.e., a bimodal distribution). To examine our procedure with bimodal distributions, we estimated thirty items for three distributions: symmetric, ceiling, and floor distributions for Likert data with small heterogeneity. Bimodal data was created by selecting half of the ceiling distribution and floor distribution to combine together. The number of bimodal items was varied from 0 percent to 100 percent increasing by 10 percent increments for simulated samples. Figure @\ref(fig:bimodal-figure) shows an example of 10 of the items within a simulation that estimated that half of the Likert items would be bimodal in nature. The researcher procedure described above was then carried out using \emph{semanticprimeR}.

\begin{verbatim}
## Warning: The dot-dot notation (`..count..`) was deprecated in ggplot2 3.4.0.
## i Please use `after_stat(count)` instead.
## This warning is displayed once every 8 hours.
## Call `lifecycle::last_lifecycle_warnings()` to see where this warning was
## generated.
\end{verbatim}

@@

Figure @\ref(fig:bimodal-figure) portrays the results of the suggested procedure. The sample with 0 bimodal items demonstrates the same values as above. The proposed sample size increases from 10 to 40\% because of the heterogeneity in SE across items (i.e., with 10\% of items with a larger SE due to their bimodal nature, the required sample size increases). From 50\% to 100\% the sample size decreases because the variance of the SEs across items decreases (i.e., at 100 percent, they are all larger rather than a mix). In theory, these results map onto the conceptual framework - if all items are truly bimodal, we have precisely measured the mean of the bimodal distribution, but this result is likely not the intended result. If researchers expect these distributions (or have representative pilot data), they could examine the data for bimodal items. These items could be estimated separately (i.e., floor effects item 1, ceiling effects item 1) to ensure that both populations of answers are represented in the data.

@@

\subsubsection{Combine Parameters}\label{combine-parameters}

@@

Second, these methods do not ensure the normal interpretation of power, focusing on finding a specific effect for a specific test, \(\alpha\), and so on. As discussed in the introduction, there is not necessarily a one-to-one mapping of hypothesis to analysis; many of the estimations within a traditional power analysis are just that - best approximations for various parameters. These proposed methods and traditional power analysis could be used together to strengthen our understanding of the sample size necessary for both a hypothesis test and a well-tuned estimation. We would advise caution when oversampling (i.e., simulation from data with larger sample sizes than the pilot data) for null-hypothesis power estimates, as these may be upwardly biased in small samples creating Type-I error (Burns et al., n.d.).

@@

\begin{itemize}
\tightlist
\item
  Data: All data used in this manuscript and vignettes have been cited and can be found on our repository pages (\url{https://github.com/SemanticPriming/stimuli-power}).
\item
  Scripts:

@@

\bibitem[\citeproctext]{ref-burns}
Burns, C. D. G., Fracasso, A., \& Rousselet, G. A. (n.d.). \emph{Data-driven estimates of the reproducibility of univariate BWAS are biased}. \url{https://doi.org/10.1101/2023.09.21.558661}

>>>>

\emph{Step 1}. The concreteness ratings data includes 27031 concepts that were rated for their concreteness. We randomly selected \emph{n} = 20 abstract words (\(M_{Rating}\) \textless= 2) and \emph{n} = 20 concrete words (\(M_{Rating}\) \textgreater= 4). In the original study, not every participant rated every word, which created uneven sample sizes for each word. Further, participants were allowed to indicate they did not know a word, and those responses were set to missing data. In our sample of 40 words, the average pilot sample size was 28.52 (\emph{SD} = 1.80), and we will use 29 as our pilot sample size for the concreteness ratings (this information will be used in the follow-up to the simulation study). We first selected the same real words in the ELP data as the concreteness subset selected above, and this data includes 27031 real words. The average pilot sample size for this random sample was 32.67 (\emph{SD} = 0.57), and \emph{n} = 33 will be our pilot size for the lexical decision task.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  How do pilot data influence sample size suggestions?
\end{enumerate}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{manuscript_draft_files/figure-latex/decile50-figure-1.pdf}}
\caption{\label{fig:decile50-figure}Comparison of the cutoffs for 5th deciles across heterogeneity (columns), powering of items (rows), and scale size (color).}
\end{figure}

@@

The above simulations assume symmetric population distributions with small, medium, or large heterogeneity or a skewed distribution with the same heterogeneity. However, as shown in Pollock (2018), sample rating data that appears to have a mean distribution may represent the average two separate distributions (i.e., a bimodal distribution). To examine our procedure with bimodal distributions, we estimated thirty items for three distributions: symmetric, ceiling, and floor distributions for Likert data with small heterogeneity. Bimodal data was created by selecting half of the ceiling distribution and floor distribution to combine together. The number of bimodal items was varied from 0 percent to 100 percent increasing by 10 percent increments for simulated samples. Figure \ref{fig:bimodal-figure} shows an example of 10 of the items within a simulation that estimated that half of the Likert items would be bimodal in nature. The researcher procedure described above was then carried out using \emph{semanticprimeR}.

@@

Figure \ref{fig:bimodal-figure} portrays the results of the suggested procedure. The sample with 0 bimodal items demonstrates the same values as above. The proposed sample size increases from 10 to 40\% because of the heterogeneity in SE across items (i.e., with 10\% of items with a larger SE due to their bimodal nature, the required sample size increases). From 50\% to 100\% the sample size decreases because the variance of the SEs across items decreases (i.e., at 100 percent, they are all larger rather than a mix). In theory, these results map onto the conceptual framework - if all items are truly bimodal, we have precisely measured the mean of the bimodal distribution, but this result is likely not the intended result. If researchers expect these distributions (or have representative pilot data), they could examine the data for bimodal items. These items could be estimated separately (i.e., floor effects item 1, ceiling effects item 1) to ensure that both populations of answers are represented in the data.

@@

\subsubsection{Combining Tools}\label{combining-tools}

@@

Second, these methods do not ensure the normal interpretation of power, focusing on finding a specific effect for a specific test, \(\alpha\), and so on. As discussed in the introduction, there is not necessarily a one-to-one mapping of hypothesis to analysis; many of the estimations within a traditional power analysis are just that - best approximations for various parameters. These proposed methods and traditional power analysis could be used together to strengthen our understanding of the sample size necessary for both a hypothesis test and a well-tuned estimation. We would advise caution when oversampling (i.e., simulation from data with larger sample sizes than the pilot data) for null-hypothesis power estimates, as these may be upwardly biased in small samples creating Type-I error (Burns et al., 2023).

@@

\begin{itemize}
\tightlist
\item
  Data: All data used in this manuscript and vignettes have been cited and can be found on our repository pages (\url{https://github.com/SemanticPriming/stimuli-power}).
\item
  Analysis Code:

@@

\bibitem[\citeproctext]{ref-burns}
Burns, C. D. G., Fracasso, A., \& Rousselet, G. A. (2023). \emph{Data-driven estimates of the reproducibility of univariate BWAS are biased}. \url{https://doi.org/10.1101/2023.09.21.558661}
